\documentclass[10pt]{article}
\usepackage[preprint]{tmlr}

% ---------- typography ----------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage{xcolor}

% ---------- math ----------
\usepackage{amsmath,amssymb,amsthm,amsfonts}
\usepackage{mathtools}
\usepackage[nodisplayskipstretch]{setspace}

% ---------- tables and lists ----------
\usepackage{booktabs}
\usepackage[shortlabels]{enumitem}
\setlist[enumerate]{topsep=2pt, partopsep=0pt, itemsep=3pt, parsep=0pt}
\setlist[itemize]{topsep=2pt, partopsep=0pt, itemsep=3pt, parsep=0pt}

% ---------- references ----------
\usepackage{hyperref}
\hypersetup{
  colorlinks,
  linkcolor={red!50!black},
  citecolor={blue!60!black},
  urlcolor={blue!80!black}
}
\usepackage{url}
\usepackage[noabbrev,capitalize]{cleveref}
\usepackage{xspace}

% ---------- airier spacing & smaller section headings ----------
\linespread{1.04}
\setlength{\parskip}{0.45em plus 0.15em minus 0.05em}
\setlength{\parindent}{0pt}
\makeatletter
% Override TMLR's oversized sectional fonts (default \large for \section).
\def\section{\@startsection{section}{1}{\z@}%
  {-2.2ex plus -0.6ex minus -.2ex}{1.4ex plus 0.3ex minus 0.2ex}%
  {\normalsize\bf\raggedright\sffamily}}
\def\subsection{\@startsection{subsection}{2}{\z@}%
  {-1.8ex plus -0.5ex minus -.2ex}{0.9ex plus .2ex}%
  {\normalsize\bf\raggedright\sffamily}}
\def\paragraph{\@startsection{paragraph}{4}{\z@}%
  {1.2ex plus 0.4ex minus .1ex}{-0.7em}{\normalsize\bf}}
\makeatother

% ----- theorem environments -----
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% ----- cleveref display names -----
\crefname{theorem}{Theorem}{Theorems}
\crefname{proposition}{Proposition}{Propositions}
\crefname{lemma}{Lemma}{Lemmas}
\crefname{corollary}{Corollary}{Corollaries}
\crefname{definition}{Definition}{Definitions}
\crefname{remark}{Remark}{Remarks}
\crefname{section}{\S}{\S\S}
\crefname{subsection}{\S}{\S\S}
\crefname{appendix}{Appendix}{Appendices}

% ----- macros -----
\newcommand{\Hbin}{H_{\mathrm{bin}}}
\newcommand{\eps}{\varepsilon}
\newcommand{\epsP}{\eps^{*}_{\Pi}}
\newcommand{\epsE}{\eps^{*}_{\varnothing}}
\newcommand{\PAMPC}{\operatorname{PA\text{-}MPC}}
\let\Pr\relax
\DeclareMathOperator{\Pr}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\LossyWL}{\mathrm{LossyWL}}
\newcommand{\Fwl}{\mathcal{F}_{\mathrm{WL}}}
\newcommand{\PiA}{\Pi_{\mathcal{A}}}
\newcommand{\dKL}{d_{\mathrm{KL}}}
\newcommand{\WL}{\mathrm{WL}}

\title{Partition-Aware Message-Passing Complexity}

\author{\name Anonymous \email anonymous@anonymous \\
        \addr Anonymous Institution}

\begin{document}
\maketitle

\begin{abstract}
It is shown that for any finite partition $\Pi$ of a finite vertex
set $V$ and any binary task $f : V \to \{0,1\}$, the cell-wise Bayes
error $\epsP$ and the partition-conditional entropy $H(f \mid \Pi)$
(in bits) satisfy the two-sided sandwich
$$
  \Hbin^{-1}\bigl(H(f \mid \Pi)\bigr) \;\le\; \epsP \;\le\; \tfrac{1}{2}\,H(f \mid \Pi).
$$
The upper bound is tight in the strong sense that no inequality
$\epsP \le c\,H(f \mid \Pi)$ with $c < 1/2$ holds uniformly.
The lower bound is tight at every $\eps \in [0, 1/2]$ and recovers the
Fano~(1961) / Hellman--Raviv~(1970) / Feder--Merhav~(1994)
entropy--error inequalities as the degenerate-partition case.

It is further shown that specialising $\Pi$ to the partition
$\PiA(G, L)$ that a depth-$L$ message-passing family $\mathcal{A}$
induces on $V(G)$ --- \emph{Partition-Aware Message-Passing
Complexity} (PA-MPC) --- yields a training-free, family-indexed
two-sided error bracket comparable across GCN, GIN, GAT, GatedGCN,
and GraphSAGE. The sandwich is verified on a finite-graph Lean~4
witness ($C_3,\ldots,C_8$ and Petersen, zero \texttt{sorry}) and on
a 1000-row exact-rational ledger. The accompanying
continuous-transfer conjecture~C1 is mathematically open; its
empirical status was frozen on 2026-05-30 as
\emph{supported-by-evidence at $L \ge 3$} on the eight-graph
non-vertex-transitive anchor.
\end{abstract}

% ============================================================
\section{Introduction}
% ============================================================

The expressivity of message-passing neural networks (MPNNs) has a
well-developed structural ceiling: the 1-Weisfeiler--Leman (WL)
discrimination power bounds what depth-$L$ MPNNs can resolve about
graphs~\citep{xu2019gin,morris2019wl}.
The partition-explicit message-passing complexity (MPC) framing
of~\citet{kemper2025mpc} refines this picture by quantifying, for the
partition a model induces, how much task information remains.

Two practical problems remain.
\begin{enumerate}
  \item \textbf{Architecture occlusion.} MPC depends on a particular
    model; its value cannot be compared meaningfully across families
    (GCN vs.\ GIN vs.\ GAT vs.\ GraphSAGE vs.\ GatedGCN) because the
    partitions those families \emph{can} induce differ at depth~0 in
    ways MPC does not expose.
  \item \textbf{No two-sided guarantee.} MPC says something about
    what the model resolves, but does not give a two-sided sandwich
    on predictability of the task by the family at that depth,
    independent of training.
\end{enumerate}

PA-MPC addresses both. By indexing on the \emph{family-level}
initial partition $\Pi^{(0)}_{\mathcal{A}}$ and applying the same
LossyWL operator to depth~$L$, PA-MPC is well-defined before any
training and is comparable across families. The bridge inequality of
\cref{sec:theory} pins its operational meaning as a two-sided
sandwich for the WL-determined task class~$\Fwl$, up to one bit.

\paragraph{Organisation and contributions.}
The rest of the paper is organised as follows.
\cref{sec:related} positions PA-MPC against three immediate
neighbours; \cref{sec:theory} develops the theory;
\cref{sec:empirical} summarises the empirical and mechanised
verification; \cref{sec:c1} reports the frozen status of
conjecture~C1; \cref{sec:limits,sec:conclusion} close the paper.
Our contributions are organised around four pillars.
\begin{enumerate}[(a)]
  \item \emph{Theory.} Definitions, the two-sided bridge inequality
    (\cref{thm:sandwich}), refinement monotonicity
    (\cref{prop:refine}), architecture factorisation
    (\cref{prop:archfact}), the PA-MPC instantiation
    (\cref{cor:pampc}), sharpness (\cref{prop:sharp}), the
    prior-aware Fano sharpening (\cref{prop:prior}), and scope.
  \item \emph{Lean witness.} A Lean~4 discrete bridge witness
    (\verb|PAMPC-LEAN-WITNESS-C4-K|) on
    $\{C_4,\ldots,C_8,\text{Petersen}\}$, zero \texttt{sorry},
    trusted-kernel-only axiom base.
  \item \emph{Empirical.} Three exact-rational ledgers (E03, E07,
    E09; tier~L-I), one float-tier oversmoothing baseline (E01),
    and three honestly-pilot results (E04, E06, E08).
  \item \emph{Reproducibility.} A sealed Merkle artefact store, a
    deterministic claim-manifest build pipeline, six decision gates,
    and methodology manuals recording exactly how every reported
    number was derived.
\end{enumerate}

\paragraph{Status discipline.}
Throughout, we keep four statuses sharply separate.
\begin{itemize}
  \item \emph{Proved (claim-pinned).} \cref{thm:sandwich},
    \cref{prop:refine,prop:archfact}, \cref{cor:pampc},
    \cref{lem:purity}.
  \item \emph{Proved (paper-only, no claim ID).}
    \cref{prop:sharp} (sharpness) and \cref{prop:prior}
    (prior-aware) are hand proofs; they are \textbf{deliberately
    not} pinned to a trust tier (L-I / L-II / float) and not part
    of the claim manifest. Their correctness rests on the appendix
    proof alone; the L-I / L-II / float infrastructure is unaffected
    by either.
  \item \emph{Mechanised / exact-audited.} Lean witness;
    exact-$\mathbb{Q}$ ledgers E02, E03, E07, E09 (L-I).
  \item \emph{Empirical / conjectural.} Float baseline E01; pilot
    apparatus E04, E06, E08; the continuous-transfer conjecture~C1
    (open; empirically frozen 2026-05-30).
\end{itemize}

% ============================================================
\section{Related Work}\label{sec:related}
% ============================================================

We position PA-MPC against three immediate neighbours. An extended
discussion of antecedents (ergodic theory, algorithmic partition
refinement, oversmoothing and over-squashing, the 2022--2024
partition-indexed GNN wave, and learning-theoretic uses of
refinement-monotonicity) is deferred to \cref{app:extrelated}.

\paragraph{Expressivity ceiling.}
\citet{xu2019gin} and \citet{morris2019wl} show that MPNNs are at
most as expressive as 1-WL on graph discrimination, and
\citet{azizian2021expressive} sharpen this to an exact
characterisation. PA-MPC is orthogonal: we \emph{quantify}
predictability of a fixed task on a fixed graph rather than
characterise isomorphism power. Higher-order $k$-WL
\citep{morris2020kwl} is explicitly out of scope; \cref{sec:scope}
gives the exclusion criterion.

\paragraph{Partition-explicit MPC.}
\citet{kemper2025mpc} frame complexity by the partition a single
trained model induces. PA-MPC lifts that framing to an
\emph{architecture-family / initial-observable pair}
$(\mathcal{A}, \Pi^{(0)})$ --- well-defined before training and
comparable across MPNN families --- and packages it with a two-sided
Bayes-error sandwich (\cref{thm:sandwich}).

\paragraph{Equitable partitions.}
The $L \to \infty$ limit of $\PiA(G, L)$ for any 1-WL-template
family is the equitable partition of $G$
\citep{schwenk1974,godsil2001ag}. That object is the most direct
intellectual ancestor of \cref{def:partition}. The novelty here is
the family indexing (\cref{prop:archfact}) and the Bayes-error
packaging (\cref{thm:sandwich}), not the partition itself.

% ============================================================
\section{Theory}\label{sec:theory}
% ============================================================

\subsection{Definitions}\label{sec:defs}

\begin{remark}[finite, measure-free setting]\label{rem:measurefree}
Throughout, $V$ is finite and we write $\Pr[\cdot] := |\{v : \cdot\}|/|V|$.
We use no $\sigma$-algebras, no countable additivity, no dominated
convergence, no Radon--Nikodým arguments.

Vertex weighting is uniform, $q_v := 1/|V|$; non-uniform weighting
amounts to replacing the counting normaliser by a positive weight
summing to $1$, with no other change in any proof.
\end{remark}

\begin{definition}[admissible partition family $\PiA$]\label{def:partition}
Let $G = (V, E)$ be a finite graph and let $\mathcal{A}$ be a
depth-$L$ MPNN architecture family with declared initial-observable
partition $\Pi^{(0)}_{\mathcal{A}}$. The \emph{depth-$\ell$ LossyWL
partition} of $\mathcal{A}$ on $G$ is
$\Pi^{(\ell)}_{\mathcal{A}} := \LossyWL^{\ell}(\Pi^{(0)}_{\mathcal{A}}; G)$
where $\LossyWL$ refines a partition by appending the multiset of
neighbour-cell labels to each vertex's label and reducing.
$\PiA(G, L) := \Pi^{(L)}_{\mathcal{A}}$.
\end{definition}

\begin{definition}[partition-conditional entropy]\label{def:pce}
For a binary task $f : V \to \{0,1\}$ and a finite partition $\Pi$
of $V$ with cells $C$, masses $q_C := |C|/|V|$, and posteriors
$P_C := \Pr[f = 1 \mid C]$,
\[
  H(f \mid \Pi) \;:=\; \sum_{C \in \Pi} q_C \cdot \Hbin(P_C)
  \;=\; \E\!\bigl[\Hbin(P_{C(v)})\bigr],
\]
where $\Hbin(p) := -p\log_2 p - (1-p)\log_2(1-p)$. Both forms
vanish iff $\Pi$ is \emph{pure} for $f$ (\cref{lem:purity}).
\end{definition}

\begin{definition}[PA-MPC]\label{def:pampc}
$\PAMPC(f, G;\, \mathcal{A}, L) := H\!\bigl(f \,\bigm|\, \PiA(G, L)\bigr).$
\end{definition}

\begin{definition}[WL-determined task class]\label{def:wldet}
$\Fwl(G)$ is the set of tasks constant on every cell of the
WL-stable partition $\Pi^{\WL}(G) := \LossyWL^{\infty}(\Pi^{(0)}; G)$.
``Determined'' rather than ``measurable'': the condition is
set-theoretic --- $f$ factors through $V \twoheadrightarrow \Pi^{\WL}$
--- and carries no measure-theoretic content.
\end{definition}

\begin{definition}[admissible architecture family]\label{def:admissible}
A pair $(\mathcal{A}, \Pi^{(0)}_{\mathcal{A}})$ is an \emph{admissible}
depth-$L$ architecture-family specification if every member
$h \in \mathcal{A}_L$ satisfies the following four conditions.
\begin{enumerate}[(1)]
  \item The depth-0 hidden state $h^{(0)}(v)$ depends on $v$ only
    through the cell of $v$ in $\Pi^{(0)}_{\mathcal{A}}$.
  \item There is a single update map $\phi_\ell$ at each layer
    $\ell$, applied identically at every vertex.
  \item $\phi_\ell$ acts on the \emph{multiset} of neighbour states.
  \item Beyond $\Pi^{(0)}_{\mathcal{A}}$, $\phi_\ell$ has no access
    to vertex names or any feature not already determined by the
    current colour class.
\end{enumerate}
GCN, GIN, GatedGCN, GAT (with permutation-invariant attention
pooling), and GraphSAGE all admit specifications meeting (1)--(4);
higher-order $k$-WL variants are out of scope (\cref{sec:scope}).
\end{definition}

\begin{lemma}[purity $\Leftrightarrow$ zero conditional entropy]\label{lem:purity}
Call $\Pi$ \emph{pure} for $f$ if $P_C \in \{0,1\}$ for every $C \in \Pi$
--- equivalently, $f$ is constant on every cell of $\Pi$, i.e.\ $f$
factors through the quotient map $V \twoheadrightarrow V/\Pi$. Then
\[
  \mathrm{Pure}(f, \Pi) \;\iff\; H(f \mid \Pi) = 0
  \;\iff\; \E\bigl[\Var(f \mid \Pi)\bigr] = 0.
\]
\end{lemma}
\begin{proof}
Each summand $q_C \cdot \Hbin(P_C)$ is non-negative and vanishes
iff $P_C \in \{0, 1\}$; the $P_C \in \{0,1\}$ condition says $f$ is
constant on $C$ (it equals its mean). The same argument applies
to $\E[\Var(f \mid \Pi)]$ via $P_C(1 - P_C) \ge 0$.
\end{proof}

\paragraph{Naming.}
The condition ``$f$ is constant on every cell of $\Pi$'' has several
established names across nearby literatures, all equivalent on a
finite vertex set:
\textbf{$\Pi$ resolves $f$} (partition-lattice combinatorics; $\ker f$
is coarsened by $\Pi$);
\textbf{$f$-equitable partition} (algebraic graph theory:
\citealp{schwenk1974,godsil2001ag});
\textbf{pure partition} (decision-tree induction:
\citealp{breiman1984cart});
\textbf{zero-equivocation partition} \citep{shannon1948}.

We use \emph{pure} as the display term --- the decision-tree mnemonic
is the closest match for a binary task --- but readers from any of
these lineages may substitute their own term verbatim; nothing in the
proofs depends on the choice.

\subsection{The Partition Bayes-Entropy Sandwich}\label{sec:sandwich}

\paragraph{Partition Bayes error.} For a finite partition $\Pi$ of
$V$ and a binary task $f$, define
\[
  \epsP \;:=\; \min_{g : \Pi \to \{0,1\}} \Pr\!\bigl[g(C(v)) \neq f(v)\bigr]
  \;=\; \sum_{C \in \Pi} q_C \cdot \min(P_C, 1 - P_C),
\]
attained by the plug-in rule $\hat{h}_{\Pi}(v) := \mathbf{1}\{P_{C(v)} \ge 1/2\}$.
Equivalently, $\epsP$ is the Bayes risk of $f$ when the predictor is
constrained to be constant on each cell of $\Pi$
\citep[\S2.1, recast for finite $V$]{devroye1996ptpr}.

\begin{theorem}[partition Bayes-entropy sandwich]\label{thm:sandwich}
Let $\Pi$ be a finite partition of a finite vertex set $V$ and let
$f : V \to \{0,1\}$ be any binary task. Then
\[
  \Hbin^{-1}\!\bigl(H(f \mid \Pi)\bigr) \;\le\; \epsP \;\le\; \tfrac{1}{2}\,H(f \mid \Pi),
\]
where $\Hbin^{-1} : [0, 1] \to [0, 1/2]$ is the inverse of binary
entropy on the increasing branch.
\end{theorem}

The theorem makes no reference to $G$, $\mathcal{A}$, the LossyWL
operator, or $\Fwl$. It is a property of any finite partition and
any binary labelling. (Upper side: \citealp{hellmanraviv1970}. Lower
side: \citealp{fano1961} in its binary form, sharpened
by~\citealp{feder1994}.) The full proof is in
\cref{app:proof}; we sketch the two halves as the two
extremal directions of a single Lagrangian.

\paragraph{Single program, two directions.} Let
$e_C := \min(P_C, 1 - P_C) \in [0, 1/2]$ and use the symmetry
$\Hbin(P_C) = \Hbin(e_C)$, so that
$H(f \mid \Pi) = \sum_C q_C \Hbin(e_C)$ and
$\epsP = \sum_C q_C e_C$. Both bounds are optimal values of opposite
Jaynes max-entropy programs on $\{(q_C, e_C)\}_C$:
\begin{align*}
\text{(Fano lower)} \quad
& \max_{\{q_C, e_C\}} \sum_C q_C \Hbin(e_C)
  \text{ s.t. } \sum_C q_C e_C = \eps
  \;\Longrightarrow\; e_C \equiv \eps, \\
\text{(HR upper)} \quad
& \max_{\{q_C, e_C\}} \sum_C q_C e_C
  \text{ s.t. } \sum_C q_C \Hbin(e_C) = H
  \;\Longrightarrow\; e_C \in \{0, 1/2\}.
\end{align*}
The Fano direction places the optimiser at the unique interior
critical point of $\Hbin$'s concavity, giving
$H \le \Hbin(\eps)$ and, on inversion,
$\Hbin^{-1}(H) \le \epsP$. The Hellman--Raviv direction places the
optimiser at the boundary of $[0, 1/2]$, where
$e \le \tfrac{1}{2}\Hbin(e)$ is tight, giving
$\epsP \le \tfrac{1}{2} H(f \mid \Pi)$.

\paragraph{Achievable region and slack.} Every finite partition and
binary task on a finite vertex set realises a pair
$(\epsP, H(f\mid\Pi))$ in
$\widetilde{A}_2 := \{(\eps, H) \in [0,1/2] \times [0,1] :
\Hbin^{-1}(H) \le \eps \le H/2\}$.
The width $w(H) := \tfrac{1}{2} H - \Hbin^{-1}(H)$ vanishes at
$H \in \{0, 1\}$ and attains its maximum
$w^{*} \approx 0.161$ at $\eps = 1/5$; the sandwich therefore
localises $\epsP$ to a window of width at most $\approx 0.161$
uniformly in $\Pi$ and $f$.

\begin{proposition}[refinement monotonicity]\label{prop:refine}
If $\Pi'$ refines $\Pi$ (every cell of $\Pi'$ is contained in some
cell of $\Pi$), then for every binary $f$,
\[
  H(f \mid \Pi') \;\le\; H(f \mid \Pi),
  \qquad
  \eps^{*}_{\Pi'} \;\le\; \epsP.
\]
\end{proposition}
\begin{proof}[Sketch]
Each $C \in \Pi$ partitions into $C = \bigsqcup_j C'_j$ with masses
$q_{C'_j}$ summing to $q_C$ and posteriors $P_{C'_j}$ averaging
(weights $q_{C'_j}/q_C$) to $P_C$. Concavity of $\Hbin$ gives
$\sum_j q_{C'_j} \Hbin(P_{C'_j}) \le q_C \Hbin(P_C)$; sum over $C$.
The same argument applies to $\min(p, 1 - p)$ via its concavity on
$[0,1]$. Full proof in \cref{app:proof}.
\end{proof}

\begin{proposition}[architecture factorisation]\label{prop:archfact}
Let $(\mathcal{A}, \Pi^{(0)}_{\mathcal{A}})$ be admissible
(\cref{def:admissible}) and let $\Pi = \PiA(G, L)$. Then
every depth-$L$ predictor $h \in \mathcal{A}_L$ is constant on
every cell of $\Pi$.
\end{proposition}
\begin{proof}
Induction on $\ell$: the layer-$\ell$ hidden state $h^{(\ell)}(v)$
depends on $v$ only through the cell of $v$ in
$\Pi^{(\ell)}_{\mathcal{A}}$. Base: \cref{def:admissible}(1).
Step: if $v, w$ share a cell of $\Pi^{(\ell+1)}_{\mathcal{A}}$ then
they share a cell of $\Pi^{(\ell)}_{\mathcal{A}}$ and the multisets
of neighbour-cell labels coincide; by induction, the multisets of
neighbour \emph{states} coincide; \cref{def:admissible}(2--4)
gives $h^{(\ell+1)}(v) = h^{(\ell+1)}(w)$.
\end{proof}

\begin{corollary}[PA-MPC sandwich for admissible families]\label{cor:pampc}
Let $(\mathcal{A}, \Pi^{(0)}_{\mathcal{A}})$ be admissible, let $G$
be a finite graph, $f : V \to \{0,1\}$ any binary task,
$\Pi = \PiA(G, L)$, and $\PAMPC = H(f \mid \Pi)$. Then:
(i) every decision rule $\widehat{F} = \tau \circ h$ with
$h \in \mathcal{A}_L$ has error $\ge \epsP \ge \Hbin^{-1}(\PAMPC)$;
(ii) the plug-in $\hat{h}_\Pi$ realises an admissible-family
predictor with error $\epsP \le \tfrac{1}{2}\PAMPC$;
(iii) increasing depth $L \to L+1$ refines $\Pi$ and so cannot
increase $\PAMPC$ or $\eps^{*}$;
(iv) if $f \in \Fwl(G)$, the bound is non-vacuous in the limit:
$\lim_{L \to \infty} \PAMPC = 0$.
\end{corollary}
\begin{proof}
(i) combines \cref{prop:archfact} with the lower bound
of \cref{thm:sandwich}; (ii) is the upper bound; (iii) is
\cref{prop:refine}; (iv) holds because
$f \in \Fwl(G)$ is constant on the WL-stable partition.
\end{proof}

\paragraph{Operational reading.}
PA-MPC sandwiches the irreducible partition Bayes error of depth-$L$
message passing on both sides --- no additive bit-slack, no
$\log$-axis ambiguity.

$\PAMPC = 0$ iff the partition is pure; positive $\PAMPC$ forces
positive $\epsP$ via the lower bound; small $\PAMPC$ forces small
$\epsP$ via the upper bound; deeper architectures never increase the
bound.

\begin{proposition}[no closed-form improvement of the sandwich]\label{prop:sharp}
Both bounds in \cref{thm:sandwich} are uniformly unimprovable in the
following sense.
\begin{enumerate}[(i)]
  \item For every constant $c < 1/2$ there exists a finite partition
    $\Pi$ and binary task $f$ on finite $V$ with
    $\epsP > c \cdot H(f \mid \Pi)$.
  \item For every \emph{continuous} $\psi : [0,1] \to [0, 1/2]$
    with $\psi(H_0) > \Hbin^{-1}(H_0)$ at some $H_0 \in (0, 1)$,
    there exist $\Pi$ and $f$ with $\epsP < \psi(H(f \mid \Pi))$.
\end{enumerate}
\end{proposition}
\begin{proof}
(i) The Hellman--Raviv witnessing family $\Pi_\alpha^{\mathrm{HR}}$
(\cref{app:proof}, Tightness) achieves
$\eps^{*}_{\Pi_\alpha^{\mathrm{HR}}} = \tfrac{1}{2} H(f \mid \Pi_\alpha^{\mathrm{HR}})$
for every $\alpha \in (0,1]$.
(ii) By continuity, $\psi(H) > \Hbin^{-1}(H)$ on a neighbourhood of
$H_0$; pick $\eps \in \mathbb{Q} \cap (0, 1/2)$ with $\Hbin(\eps)$ in
that neighbourhood and apply the Fano witnessing family at any $n$
with $\eps n \in \mathbb{Z}$.
\end{proof}

The rhetorical content of \cref{prop:sharp} is
Massey-style \citep[\S III]{massey1994}: a positive sharpness claim
upgraded to a negative non-existence claim. The achievable region
$\widetilde{A}_2$ is the tightest two-sided closed-form bracket on
$(\epsP, H(f \mid \Pi))$ available in the binary case.

\begin{proposition}[prior-aware sharpening]\label{prop:prior}
Let $\Pi$ be a finite partition of finite $V$ and $f : V \to \{0,1\}$
a binary task. With marginal frequency
$P_f := |\{v : f(v) = 1\}| / |V|$, marginal Bayes error
$\epsE := \min(P_f, 1 - P_f)$, and mutual information
$I(f ; \Pi) := \Hbin(P_f) - H(f \mid \Pi)$,
\[
  \dKL\!\bigl(\epsP \,\bigm\|\, \epsE\bigr) \;\le\; I(f ; \Pi),
\]
where $\dKL(p \,\|\, q) := p \log_2 \tfrac{p}{q} + (1-p) \log_2 \tfrac{1-p}{1-q}$.
\end{proposition}
\begin{proof}
\cref{app:proof}, equations (A.10)--(A.12), via
data-processing on the binary error indicator \citep{hanverdu1994}.
\end{proof}

\paragraph{Status (trust-tier note).}
\cref{prop:prior} is a \textbf{paper-only hand proof}.
It carries no \verb|PAMPC-*| claim ID, is not part of the L-I
exact-rational ledger, not part of the L-II Lean witness, and not
behind any G0..G4 decision gate. The L-I / L-II / float trust-tier
infrastructure neither audits nor depends on it, and a future flaw
in this proposition would not invalidate any claim-pinned result.
Substituting the trivial bound $\epsE = 1/2$ in
\cref{prop:prior} recovers exactly
\cref{thm:sandwich}'s lower bound; for skewed marginals the
bound is strictly tighter.

\subsection{Scope}\label{sec:scope}

The partition theorem (\cref{thm:sandwich}) and refinement
proposition (\cref{prop:refine}) are stated for \emph{any} finite
partition and \emph{any} binary task on a finite vertex set; nothing
about graphs, WL, or admissible architectures enters those
statements.

The architecture factorisation (\cref{prop:archfact}) and the PA-MPC
sandwich (\cref{cor:pampc}) require the admissibility conditions of
\cref{def:admissible}. The non-vacuity clause \cref{cor:pampc}(iv) is
the one place where $f \in \Fwl(G)$ is used.

Higher-order WL templates \citep{morris2020kwl}, which violate
admissibility~(4) by referencing $k$-tuples, are out of scope.
Multi-class extensions $(|\mathcal{Y}| > 2)$ are routine via the
concave-envelope construction of~\citet{feder1994} and are deferred
to follow-up work.

\subsection{Three named conjectures (registry)}\label{sec:conj}

\begin{table}[h]
\centering
\caption{Conjecture registry. C1 is the only conjecture whose
resolution gates a phase transition in the programme; its precise
statement, hypotheses H1--H3, decision rule, and sealed empirical
verdict are in \cref{sec:c1}.}
\label{tab:conj}
\begin{tabular}{@{}lll@{}}
\toprule
ID & Title & Falsifier \\
\midrule
C1 & Continuous-transfer (open; frozen 2026-05-30) & E08 (\cref{sec:c1}) \\
C2 & $K_B$ rank saturation on cycles & E03 \\
C3 & Feature-richness boundary & E07 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Remark on $\mathrm{MI}^2 \approx 1/2$}\label{sec:mi2}

Under purity, $\mathrm{MI}^2(f ; \Pi) = \tfrac{1}{2} R^2(f ; \Pi) = \tfrac{1}{2}$
regardless of cell count --- operationally vacuous in the solvable
regime. The partition-conditional entropy is the \emph{informative
dual}: where $\mathrm{MI}^2$ becomes constant, the conditional
entropy vanishes; before that point, it carries the
architecture-level signal.

% ============================================================
\section{Empirical and Mechanised Verification (summary)}\label{sec:empirical}
% ============================================================

The full experimental record (E01--E09, per-experiment LAB-PLANs,
exact-rational ledger contents, Lean module listings, and the
reproducibility infrastructure) is held in the project's Markdown
companion (\verb|PAPER-ARXIV.md|, \S4--\S8) and the
\verb|20-experiments/| tree. We summarise the headline outcomes.

\paragraph{Exact-rational ledger (tier L-I).} Three sealed ledgers
verify the theory on a synthetic anchor with L-I-computable WL
partition and exact-$\mathbb{Q}$ partition-conditional entropy.
\textbf{E03} (audit of \cref{prop:refine}): $280$ rows
($\{P_4,\ldots,P_{12},\text{Petersen}\} \times 4$ tasks $\times 7$
depths); all three falsifiers PASS (monotonicity, purity on
vertex-transitive, WL-stability scaling).
\textbf{E07} (feature-richness boundary): GIN- vs.\ GCN-archetype
init on the same grid; refinement chain holds on $280/280$ rows,
$18$ strict-improvement rows witness Conjecture C3 non-vacuous, zero
regressions.
\textbf{E09} (intervention pricing): rewiring price $\ge 0$ on WL-cell
tasks, all invariants PASS; sign-variability is preserved as an
honest non-gating advisory.

\paragraph{Lean witness (tier L-II).} A Lean~4 module verifies the
discrete bridge on $\{C_4, \ldots, C_8, \text{Petersen}\}$ at depths
$0..3$ via \verb|native_decide|; zero \texttt{sorry};
trusted-kernel-only axiom base
(\verb|Classical.choice|, \verb|propext|).
Claim \verb|PAMPC-LEAN-WITNESS-C4-K|.

\paragraph{Float-tier baseline.} \textbf{E01} (oversmoothing): lazy
random walk on bipartite anchors reproduces depth-induced collapse
qualitatively (smoothed-window F2). \textbf{E04, E06, E08} are
honestly pilot.

\paragraph{Reproducibility skeleton.} Sealed Merkle artefact store
(10 experiments / 27 leaves); deterministic claim-manifest build
pipeline; six decision gates (G0..G4 PASS).

% ============================================================
\section{Continuous-Transfer Conjecture C1 (empirical status)}\label{sec:c1}
% ============================================================

\textbf{C1 (continuous-transfer; mathematically open).} For every
admissible MPNN family $\mathcal{A}$ and every binary task
$f \in \Fwl(G)$, the depth-monotone PA-MPC trajectory
$L \mapsto \PAMPC(f, G; \mathcal{A}, L)$ admits a continuous
$L$-extension on a sealed substrate, with stage-resolved decay rates
that depend only on the architectural class
$\{\text{GCN}, \text{GIN}, \text{GatedGCN}\}$ vs.\
$\{\text{GAT}, \text{GraphSAGE}\}$.

\paragraph{Empirical verdict (frozen 2026-05-30).} Under the
literal \S9 protocol of \verb|PAPER-ARXIV.md|, status
\emph{supported-by-evidence at $L \ge 3$} on the eight-graph
non-vertex-transitive anchor $\mathrm{anchor}_8$, with the
architectural dichotomy
$\{\text{GCN}, \text{GIN}, \text{GatedGCN}\} \,/\, \{\text{GAT}, \text{GraphSAGE}\}$.
The companion conjecture C1' (2-WL substrate, E10) is the headline
open problem for paper-02.

% ============================================================
\section{Known Limitations and Honest Narrowings}\label{sec:limits}
% ============================================================

\begin{enumerate}
\item \textbf{E06 demoted} from headline to inconclusive-pilot.
\item \textbf{E09 F3 reformulated} from sign-variability to
  ``rewiring price $\ge 0$ on WL-cell tasks''.
\item \textbf{E01 required three task/operator redesigns.}
\item \textbf{G4 narrowed} from PDF to claim manifest
  (byte-determinism without strict \verb|SOURCE_DATE_EPOCH|).
\item \textbf{No \texttt{mode=full} for E04 and E08.} Sealed
  Stage~K/L/M C1 verdict only.
\item \textbf{No real-world graph results.} ZINC/OGB/TUDataset are
  in scope for paper-02; this paper is theory plus the synthetic
  anchor.
\item \textbf{Propositions~\ref{prop:sharp} and~\ref{prop:prior}
  are paper-only.} Both are hand proofs in
  Appendix~\ref{app:proof} and \textbf{by design} carry no
  \verb|PAMPC-*| claim ID, no trust tier, and no decision gate.
  Lean mechanisation and an E02 demo row for
  Proposition~\ref{prop:prior} are listed as \emph{optional}
  future work; the L-I / L-II / float infrastructure is unaffected
  by either.
\end{enumerate}

% ============================================================
\section{Conclusion}\label{sec:conclusion}
% ============================================================

PA-MPC is a partition-explicit, architecture-family-indexed
generalisation of MPC. The discrete bridge inequality
(\cref{thm:sandwich}) pins its operational meaning as a two-sided
sandwich on WL-determined predictability. The exact-rational ledger
and the Lean L-II witness verify the theory on a synthetic anchor.

The continuous-transfer conjecture~C1 is frozen as of 2026-05-30:
supported-by-evidence at $L \ge 3$ on $\mathrm{anchor}_8$, with the
architectural dichotomy
$\{\text{GCN}, \text{GIN}, \text{GatedGCN}\}$ versus
$\{\text{GAT}, \text{GraphSAGE}\}$. The companion conjecture~C1' is
registered as the headline open problem for paper-02.

The reproducibility skeleton is closed end-to-end. The kill branch
of C1 is publishable; the PASS branch is publishable; and the honest
pilot status of E04, E06, E08 is publishable.

\bibliographystyle{tmlr}
\bibliography{main}

\newpage
\appendix

% ============================================================
\section{Proof of Theorem~\ref{thm:sandwich}}\label{app:proof}
% ============================================================

\textbf{Setup.} Fix a finite vertex set $V$, a binary task
$f : V \to \{0, 1\}$, and a finite partition $\Pi$ of $V$ with
cells $C$, masses $q_C := |C|/|V|$, and posteriors
$P_C := \Pr[f = 1 \mid C]$. By Remark~\ref{rem:measurefree}, vertex
weighting is uniform, $q_v := 1/|V|$.

\paragraph{Upper bound (Hellman--Raviv).} For each cell, let
$e_C := \min(P_C, 1 - P_C) \in [0, 1/2]$. The pointwise inequality
$e \le \tfrac{1}{2} \Hbin(e)$ on $[0, 1/2]$ \citep{hellmanraviv1970}
gives, summing weighted by $q_C$ and using the symmetry
$\Hbin(P_C) = \Hbin(e_C)$,
\[
  \epsP \;=\; \sum_C q_C e_C \;\le\; \tfrac{1}{2} \sum_C q_C \Hbin(e_C) \;=\; \tfrac{1}{2} H(f \mid \Pi).
  \tag{A.1--A.2}
\]

\paragraph{Lower bound (Fano, binary form).} The function $\Hbin$ is
concave on $[0, 1/2]$. Jensen's inequality applied to the cell-mass
distribution $\{q_C\}$ and the values $\{e_C\}$ yields
\[
  H(f \mid \Pi) \;=\; \sum_C q_C \Hbin(e_C) \;\le\; \Hbin\!\bigl(\textstyle\sum_C q_C e_C\bigr) \;=\; \Hbin(\epsP).
  \tag{A.3--A.5}
\]
Since $\Hbin$ is monotone-increasing on $[0, 1/2]$ with
inverse $\Hbin^{-1} : [0, 1] \to [0, 1/2]$,
\[
  \Hbin^{-1}\!\bigl(H(f \mid \Pi)\bigr) \;\le\; \epsP.
  \tag{A.6}
\]
Combining (A.2) and (A.6) gives Theorem~\ref{thm:sandwich}.

\paragraph{Proof of Proposition~\ref{prop:refine}.} If $\Pi'$ refines
$\Pi$, each $C \in \Pi$ partitions as
$C = \bigsqcup_j C'_j$ with $q_{C'_j}$ summing to $q_C$. The cell
posterior $P_C$ equals the $q_{C'_j}$-weighted average of $\{P_{C'_j}\}_j$,
so by concavity of $\Hbin$,
$\sum_j q_{C'_j} \Hbin(P_{C'_j}) \le q_C \Hbin(P_C)$. Summing over
$C$ gives $H(f \mid \Pi') \le H(f \mid \Pi)$. The same argument
with the concave $p \mapsto \min(p, 1-p)$ on $[0,1]$ gives
$\eps^{*}_{\Pi'} \le \epsP$. \qed

\paragraph{Proof of Proposition~\ref{prop:prior} (prior-aware sharpening).}
Let $P_f := |\{v : f(v) = 1\}| / |V|$ and
$\epsE := \min(P_f, 1 - P_f)$. Define the binary error indicator
$Z(v) := \mathbf{1}\{\hat{h}_\Pi(v) \neq f(v)\}$ for the plug-in
rule $\hat{h}_\Pi$ of \S\ref{sec:sandwich}. By construction
$\Pr[Z = 1] = \epsP$. Since $Z$ is a deterministic function of
$(f, \Pi)$, the finite-alphabet data-processing inequality
\citep[Thm 2.8.1, specialised to discrete distributions --- no
measure-theoretic generality needed]{cover2006eit} on the Markov
chain $f \to \Pi \to Z$ gives
\[
  I(f ; Z) \;\le\; I(f ; \Pi).
  \tag{A.10}
\]
The joint distribution of $(f, Z)$ has marginals $P_f$ and $\epsP$
and is fully determined by $\Pr[f \neq \hat{h}_\Pi] = \epsP$. Among
all couplings of $(f, Z)$ with these marginals, the minimum mutual
information is attained when $\hat{h}_\Pi$ predicts the marginal
mode and makes errors only on the minority; this minimum equals
$\dKL(\epsP \| \epsE)$ \citep[eq.\ (6)]{hanverdu1994}. Hence
\[
  I(f ; Z) \;\ge\; \dKL\!\bigl(\epsP \,\bigm\|\, \epsE\bigr).
  \tag{A.11}
\]
Combining (A.10)--(A.11),
\[
  \dKL\!\bigl(\epsP \,\bigm\|\, \epsE\bigr) \;\le\; I(f ; \Pi).
  \tag{A.12}
\]
Theorem~\ref{thm:sandwich}'s lower bound is the special case
$P_f = 1/2$ (then $\epsE = 1/2$ and
$\dKL(\eps \| 1/2) = 1 - \Hbin(\eps)$, so (A.12) reduces to
$H(f \mid \Pi) \le \Hbin(\epsP)$, equation (A.6)). For $P_f \neq 1/2$
the bound is strictly tighter. \qed

\paragraph{Tightness.} Both boundaries of $\widetilde{A}_2$ are
realised by explicit one-parameter families on finite $V$ (uniform
vertex weighting, Remark~\ref{rem:measurefree}).
\emph{Fano (lower) boundary $\Pi_\eps^{\mathrm{F}}$:} for
$\eps \in [0, 1/2]$ and $n$ with $\eps n \in \mathbb{Z}$, take
$\Pi_\eps^{\mathrm{F}} := \{V\}$ and let $f$ have exactly $\eps n$
ones; then $P = \eps = e$ and Jensen is tight.
\emph{Hellman--Raviv (upper) boundary $\Pi_\alpha^{\mathrm{HR}}$:}
take two cells of equal mass with $(P_1, P_2) = (1/2, 1/2 - \alpha)$
so that $e_1 = 1/2$, $e_2 = \alpha$ and the pointwise inequality
$e \le \tfrac{1}{2}\Hbin(e)$ is tight at both endpoints.

% ============================================================
\section{Extended Related Work}\label{app:extrelated}
% ============================================================

This appendix collects context that \S\ref{sec:related} only
gestures at. None of it is load-bearing for the proofs or for the
empirical claims.

\paragraph{Oversmoothing and over-squashing.} \citet{cai2020oversmooth}
and \citet{oono2020graph} characterise depth-induced embedding
collapse on bipartite graphs; our float-tier experiment E01
reproduces this qualitatively. \citet{alon2021bottleneck} frame
over-squashing as an information-bottleneck phenomenon. The bridge
inequality is agnostic to bottleneck architecture: it holds for any
MPNN computing via depth-$L$ rooted coloured neighbourhood
aggregation.

\paragraph{Partition-refinement antecedents.}
\textit{(i) Ergodic theory.} \citet{kolmogorov1958},
\citet{sinai1959}, and \citet{rokhlin1967} defined entropy as a
supremum over refining partitions; refinement-monotonicity of
$H(\Pi)$ is the structural axiom of that construction.
Proposition~\ref{prop:refine} is the finite, combinatorial cousin
(no $\sigma$-additivity used --- see Remark~\ref{rem:measurefree}).
\textit{(ii) Algorithmic partition refinement.} \citet{hopcroft1971}
and \citet{paige1987} gave the algorithmic schema; 1-WL is the
graph specialisation, used in McKay's \emph{nauty}~\citep{mckay1981}
for isomorphism testing and sharpened into the $k$-WL hierarchy
by \citet{caifurer1992} and \citet{grohe2017}.
\textit{(iii) Algebraic graph theory.} The equitable-partition
ancestry is already foregrounded in \S\ref{sec:related};
\citet{schwenk1974} and \citet[ch.~9]{godsil2001ag} are the direct
sources for the $L \to \infty$ fixpoint statement.

\paragraph{Modern partition-indexed GNN theory (2022--2024).}
\citet{geerts2022} characterise MPNN expressivity by per-layer
partition refinement; \citet{morris2023survey} survey the field
with equitable partitions as the WL fixpoint;
\citet{grohe2021logic} connects partition refinement to
counting-logic; \citet{boker2024finegrained} define a metric on
partitions and quantify how close MPNNs come to separating points
--- the closest existing work in spirit to our Bayes-error
packaging. None of this wave delivers all three of (a) a
two-sided Bayes-error bracket indexed by $\PiA(G, L)$, (b) an
exact-rational ledger on small graphs, and (c) a mechanised
statement; that triple is the unique contribution of PA-MPC.

\paragraph{Learning-theoretic uses of refinement-monotonicity.}
Outside GNNs, \citet{steinke2020chaining} and
\citet{asadi2018chaining} apply refinement-monotonicity arguments
to partitions of hypothesis classes for generalisation bounds ---
the same Proposition~\ref{prop:refine} move applied in learning
theory.

\end{document}
