% =============================================================
% Paper B — Partition Brackets: A Framework with Entropy,
% Variance, and Noise-Robust Instances
%
% STATUS: skeleton (Phase 2a scaffold). All numbered statements
% are placeholders; proofs are marked "skeleton" and will be
% completed in Phase 2b (theory) and Phase 2c (instances).
% Do NOT cite or distribute until Gate G2 closes.
% =============================================================
\documentclass[10pt]{article}
\usepackage[preprint]{tmlr}

% ---------- typography ----------
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{microtype}
\usepackage{xcolor}

% ---------- math ----------
\usepackage{amsmath,amssymb,amsthm,amsfonts}
\usepackage{mathtools}

% ---------- lists ----------
\usepackage[shortlabels]{enumitem}

% ---------- figures & tables ----------
\usepackage{graphicx}
\usepackage{booktabs}

% ---------- references ----------
\usepackage{hyperref}
\hypersetup{
  colorlinks,
  linkcolor={red!50!black},
  citecolor={blue!60!black},
  urlcolor={blue!80!black}
}
\usepackage[noabbrev,capitalize]{cleveref}

% ----- theorem environments -----
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}
\newtheorem{conjecture}[theorem]{Conjecture}

\crefname{theorem}{Theorem}{Theorems}
\crefname{proposition}{Proposition}{Propositions}
\crefname{lemma}{Lemma}{Lemmas}
\crefname{corollary}{Corollary}{Corollaries}
\crefname{remark}{Remark}{Remarks}
\crefname{definition}{Definition}{Definitions}
\crefname{example}{Example}{Examples}
\crefname{conjecture}{Conjecture}{Conjectures}
\crefname{figure}{Figure}{Figures}
\crefname{section}{\S}{\S\S}

% ----- macros (shared with Paper A) -----
\newcommand{\Hbin}{H_{\mathrm{bin}}}
\newcommand{\Hinv}{H_{\mathrm{bin}}^{-1}}
\newcommand{\eps}{\varepsilon}
\newcommand{\epsP}{\eps^{\!*}_{\Pi}}
\newcommand{\epsK}{\eps^{\!*}_{K}}      % soft / Markov-kernel risk
% versions that take a loss superscript (avoids double-superscript)
\newcommand{\epsPL}[1]{\eps^{*,#1}_{\Pi}}
\newcommand{\epsKL}[1]{\eps^{*,#1}_{K}}
\let\Pr\relax
\DeclareMathOperator{\Pr}{\mathbb{P}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\KL}{D_{\mathrm{KL}}}
\newcommand{\TV}{\lVert\cdot\rVert_{\mathrm{TV}}}
\newcommand{\bphi}{\bar{\varphi}}        % concave score functional
\newcommand{\paperA}{Paper~A}

% status macro for skeleton claims
\newcommand{\status}[1]{\par\noindent\emph{Status:} #1}

\title{Partition Brackets: A Framework with\\
       Entropy, Variance, and Noise-Robust Instances}

\author{\name Anonymous \email anonymous@anonymous \\
        \addr Anonymous Institution}

\begin{document}
\defcitealias{paperA}{Paper~A}
\maketitle

\begin{abstract}
The companion paper \citep{paperA} proves an elementary two-sided
closed-form bracket between the partition-restricted Bayes risk
$\epsP$ of a binary label $f$ under a partition $\Pi$ and its
partition-conditional Shannon entropy $H(f\mid\Pi)$. The present
paper argues that this bracket is one instance of a single
\emph{recipe}: pick any concave \emph{score functional} $\bphi$ of
a label distribution, take its partition-conditional expectation,
and read off a two-sided bracket on the corresponding partition-
restricted Bayes loss. We formalise the recipe as a $\bphi$-bracket
\emph{meta-theorem}, instantiate it on three concrete score
functionals (Shannon entropy, conditional variance, Pinsker /
KL), and prove robustness extensions in three directions:
(i) symmetric label noise via population correction, (ii) soft /
Markov-kernel partitions where cell assignment is itself a
distribution, and (iii) refinement chains where one partition
refines another. The framework recovers \citetalias{paperA}'s
binary-entropy result as the Shannon instance, recovers the
classical Bayes--variance identity as the variance instance, and
yields a new MAE / regression instance via Cauchy--Schwarz. We
illustrate the framework on a quantised neural-network
``partition-as-bottleneck'' demonstration and on a label-noise
stress test, both of which rely on \emph{zero new training}
beyond \citetalias{paperA}'s released model artefacts.
\end{abstract}

% ============================================================
\section{Introduction}\label{sec:intro}
% ============================================================

\paragraph{Companion paper.} This paper is a direct sequel to
\citet{paperA}. We assume the reader is familiar with the
partition framework introduced there: a partition
$\Pi=\{S_1,\dots,S_m\}$ of the input space $\mathcal{X}$, the
\emph{partition-restricted Bayes risk}
$\epsP := \min_{g\in G(\Pi)} \E[\mathbf{1}\{g(X)\ne f(X)\}]$
where $G(\Pi)$ is the class of predictors constant on every
cell, and the partition-conditional entropy
$H(f\mid\Pi)=\sum_i p_i\,\Hbin(\eta_i)$ with cell mass
$p_i=\Pr(\Pi(X)=S_i)$ and cell-conditional positive rate
$\eta_i=\Pr(f(X)=1\mid \Pi(X)=S_i)$.

\paragraph{Thesis.} \citetalias{paperA}'s two-sided bracket
\[
  \Hinv\!\bigl(H(f\mid\Pi)\bigr)
  \;\le\; \epsP
  \;\le\; \tfrac12\,H(f\mid\Pi)
\]
is not an isolated coincidence of binary entropy. We claim it is
the Shannon instance of a one-line recipe that holds for every
concave score functional $\bphi$ with $\bphi(0)=\bphi(1)=0$ and a
matching pointwise loss $\ell_{\bphi}$. The recipe gives a
two-sided bracket on $\epsPL{\ell_{\bphi}}$, the
partition-restricted Bayes loss for $\ell_{\bphi}$, expressed
through the partition-conditional functional
$\bphi(f\mid\Pi):=\sum_i p_i\,\bphi(\eta_i)$.

\paragraph{Contributions.}
\begin{itemize}[leftmargin=*,nosep]
  \item A $\bphi$-bracket \emph{meta-theorem}
    (\cref{thm:phi-bracket}) generalising \citetalias{paperA}'s
    main theorem.
  \item Three named instances of the meta-theorem
    (\cref{sec:instances}): Shannon entropy (recovers
    \citetalias{paperA}), conditional variance (Bayes--variance
    identity), and the Pinsker / KL instance.
  \item A regression instance for MSE (identity) and MAE
    (Cauchy--Schwarz upper bound), \cref{thm:regression}.
  \item Population-exact correction for symmetric label noise,
    \cref{thm:noise}, extending \citetalias{paperA}'s
    \texttt{Proposition~7} to arbitrary $\bphi$.
  \item Soft / Markov-kernel partitions, \cref{thm:soft}, with
    explicit definition of $\epsK$ and bracket re-derivation.
  \item Aggregator-typed Lipschitz constants for MPNN
    Weisfeiler--Leman robustness (sum / mean / sym-norm),
    \cref{lem:mpnn-typed}, in the $\bphi$-bracket form.
  \item Two experiments using \citetalias{paperA}'s released
    artefacts: E8 (variance vs.\ entropy bracket across all Paper
    A partitions) and E9 (label-noise stress test, validates
    \cref{thm:noise}). Both reuse model checkpoints; zero new
    training.
\end{itemize}

\paragraph{Relation to \citetalias{paperA}.}
\citetalias{paperA}'s purpose is to ship one elementary,
training-free, closed-form diagnostic for ML practitioners. The
present paper's purpose is to expose the structural reason
\citetalias{paperA} works and to advertise the recipe. The two
papers therefore have intentionally disjoint audiences and
disjoint emphases: \citetalias{paperA} is a TMLR /
\emph{methods} paper; the present paper is a COLT / TMLR-long
\emph{theory} paper. We pay an explicit citation tax to
\citetalias{paperA} for the binary-entropy bracket; we do not
re-prove it.

% ============================================================
\section{The \texorpdfstring{$\bphi$}{phi}-bracket
  meta-theorem}\label{sec:meta}
% ============================================================

\begin{definition}[Concave score functional]\label{def:score}
A function $\bphi:[0,1]\to[0,\infty)$ is a \emph{concave score
functional} if it is concave, continuous, and satisfies
$\bphi(0)=\bphi(1)=0$. Its \emph{matched pointwise loss}
$\ell_{\bphi}:\{0,1\}\times[0,1]\to[0,\infty)$ is the unique
pointwise Bayes-consistent loss whose Bayes-optimal predictor on
a Bernoulli$(\eta)$ label has expected loss $\bphi(\eta)$; for
the three named cases this is:
\begin{itemize}[leftmargin=*,nosep]
  \item $\bphi=\Hbin$, $\ell_{\bphi}(y,\hat p)=-y\log\hat p
        -(1-y)\log(1-\hat p)$ (log-loss).
  \item $\bphi(\eta)=\eta(1-\eta)$,
        $\ell_{\bphi}(y,\hat p)=(y-\hat p)^2$ (squared loss).
  \item $\bphi(\eta)=\min(\eta,1-\eta)$, $\ell_{\bphi}=$
        $0$--$1$ loss (the recipe collapses to
        \citetalias{paperA}'s setting upon binarising).
\end{itemize}
\end{definition}

\begin{theorem}[$\bphi$-bracket meta-theorem]\label{thm:phi-bracket}
Let $\Pi$ be a partition of $\mathcal{X}$, let $f:\mathcal{X}\to
\{0,1\}$ be measurable, and let $\bphi$ be a concave score
functional (\cref{def:score}) with matched loss $\ell_{\bphi}$.
Let
$\bphi(f\mid\Pi):=\sum_i p_i\,\bphi(\eta_i)$
denote the partition-conditional functional, and let
$\epsPL{\ell_{\bphi}}:=\min_{g\in G(\Pi)}\E\,\ell_{\bphi}(f(X),g(X))$
denote the partition-restricted Bayes loss for $\ell_{\bphi}$.
Then
\[
  \underbrace{\bphi^{-1}\!\bigl(\bphi(f\mid\Pi)\bigr)}_{\text{Jensen lower}}
  \;\le\;
  \epsPL{\ell_{\bphi}}
  \;\le\;
  \underbrace{c_{\bphi}\,\bphi(f\mid\Pi)}_{\text{linear upper}},
\]
where $\bphi^{-1}$ is the restriction of $\bphi$ to
$[0,\tfrac12]$ inverted, and the upper constant
$c_{\bphi}:=\sup_{\eta\in[0,1]}
\ell_{\bphi}^{*}(\eta)/\bphi(\eta)$ is finite for all three
named instances. The Shannon instance recovers
\citetalias{paperA}'s bracket with
$c_{\Hbin}=\tfrac12$.
\status{Skeleton. Full proof in Phase~2b; explicit hypothesis
list and Jensen-sharp lower bound construction pending.}
\end{theorem}

% ============================================================
\section{Three named instances}\label{sec:instances}
% ============================================================

\begin{corollary}[Shannon instance: \citetalias{paperA}]
\label{cor:shannon}
With $\bphi=\Hbin$ and $\ell_{\bphi}=$ log-loss, \cref{thm:phi-bracket}
specialises to
$\Hinv(H(f\mid\Pi))\le \epsP\le \tfrac12 H(f\mid\Pi)$
(under the $0$--$1$ collapse of log-loss to its Bayes
classifier). The upper constant $c_{\Hbin}=\tfrac12$ is sharp;
see \citetalias{paperA}.
\status{Reduction proof in Phase~2b.}
\end{corollary}

\begin{corollary}[Variance / squared-loss instance]
\label{cor:variance}
With $\bphi(\eta)=\eta(1-\eta)$ and $\ell_{\bphi}=$ squared
loss, \cref{thm:phi-bracket} recovers the classical
Bayes--variance identity:
\[
  \epsPL{\mathrm{MSE}} \;=\; \E[\,\Var(f\mid\Pi(X))\,]
  \;=\; \bphi(f\mid\Pi),
\]
so the bracket collapses to equality on both sides.
\status{Identity well known; serves as the meta-theorem's
sanity check.}
\end{corollary}

\begin{corollary}[Pinsker / KL instance]\label{cor:pinsker}
With $\bphi(\eta)=\KL(\eta\,\|\,\tfrac12)$ and log-loss,
\cref{thm:phi-bracket} yields a non-trivial bracket sharper
than Shannon when cell purities concentrate near
$0$ or $1$; the upper constant is
$c_{\KL}\le 1/(2\ln 2)$ via Pinsker's inequality.
\status{Hypothesis-checking pending Phase~2b.}
\end{corollary}

\begin{theorem}[Regression instance]\label{thm:regression}
Let $f:\mathcal{X}\to\mathbb{R}$ be square-integrable. With
$\bphi(\eta)=\Var(f\mid\eta)$ (cell-conditional variance), the
meta-theorem yields:
\begin{itemize}[leftmargin=*,nosep]
  \item \emph{MSE identity:} $\epsPL{\mathrm{MSE}}=\E[\Var(f\mid\Pi(X))]$.
  \item \emph{MAE upper bound via Cauchy--Schwarz:}
        $\epsPL{\mathrm{MAE}}\le \sqrt{\,\E[\Var(f\mid\Pi(X))]\,}$.
\end{itemize}
\status{MSE identity is folklore. MAE direction reformulated
from \citetalias{paperA}'s T7 audit; matching lower bound
(Hotelling 1932 path) is an open problem flagged in Phase~2.2.}
\end{theorem}

% ============================================================
\section{Robustness: noise, soft partitions, refinement}
\label{sec:robust}
% ============================================================

\begin{theorem}[Symmetric label-noise correction]\label{thm:noise}
Let $\tilde f$ be the label $f$ corrupted by symmetric
flip-noise of rate $\eta_{\mathrm{noise}}<\tfrac12$. Then for
every concave score functional $\bphi$ admitting a noise-corrected
analogue $\bphi_{\eta_{\mathrm{noise}}}$,
\[
  \bphi^{-1}\!\bigl(\bphi_{\eta_{\mathrm{noise}}}(\tilde f\mid\Pi)\bigr)
  \;\le\;
  \epsPL{\ell_{\bphi}}(f)
  \;\le\;
  c_{\bphi}\,\bphi_{\eta_{\mathrm{noise}}}(\tilde f\mid\Pi).
\]
The Shannon instance recovers \citetalias{paperA}'s
population-exact correction. The variance instance gives the
classical noise-corrected Bayes--variance identity.
\status{Skeleton; full proof in Phase~2b. Sharpness witness
TBD.}
\end{theorem}

\begin{definition}[Markov-kernel partition]\label{def:markov}
A \emph{Markov-kernel partition} is a stochastic kernel
$K:\mathcal{X}\to\Delta(\{1,\dots,m\})$ whose cell-assignment
is itself a probability distribution over $m$ soft cells. The
\emph{soft partition-restricted Bayes loss} $\epsKL{\ell_{\bphi}}$
is the loss of the best predictor that is a function of $K(X)$
only.
\end{definition}

\begin{theorem}[Soft / Markov-kernel bracket]\label{thm:soft}
Under \cref{def:markov}, the $\bphi$-bracket
\cref{thm:phi-bracket} holds with $\Pi$ replaced by $K$ and the
cell-conditional functional
$\bphi(f\mid K):=\E_{X}\,\bphi(\Pr(f=1\mid K(X)))$. The hard
partition is the deterministic-kernel special case.
\status{Skeleton; full proof in Phase~2b. Instantiation on
attention soft-routing in Phase~2c.}
\end{theorem}

\begin{proposition}[Refinement consistency, $\bphi$-version]
\label{prop:refine-phi}
If $\Pi'$ refines $\Pi$, then $\bphi(f\mid\Pi')\le\bphi(f\mid\Pi)$
and consequently both sides of the $\bphi$-bracket are
monotone in refinement. The discrete limit
($\Pi'=\{\{x\}:x\in\mathcal{X}\}$) pinches the bracket to the
true Bayes loss whenever $\bphi$ is non-degenerate.
\status{Mirrors \citetalias{paperA}'s Proposition~4.5;
specialise its proof to general $\bphi$ in Phase~2b.}
\end{proposition}

% ============================================================
\section{Aggregator-typed MPNN Lipschitz, $\bphi$-form}
\label{sec:mpnn}
% ============================================================

\begin{lemma}[Aggregator-typed Lipschitz, $\bphi$-version]
\label{lem:mpnn-typed}
Let an MPNN of depth $L$ have per-layer combine-Lipschitz $L_\ell^c$
and message-Lipschitz $L_\ell^m$. With aggregator-dependent
amplification factor $r_T\in\{\Delta,1,1\}$ for
sum / mean / sym-norm respectively, the $\bphi$-bracket
\cref{thm:phi-bracket} for the WL-induced partition $\Pi_{\mathrm{WL}}^L$
holds with $\bphi$-slack at most
$\delta_L=\delta_0\,\prod_{\ell=1}^L(L_\ell^c+r_T L_\ell^m)$.
\status{Direct $\bphi$-lift of \citetalias{paperA}'s
Lemma~6$'$; proof to be transcribed in Phase~2b.}
\end{lemma}

% ============================================================
\section{Experiments}\label{sec:exp}
% ============================================================

We describe two experiments that reuse \citetalias{paperA}'s
released artefacts and require \emph{zero} new training.
\begin{itemize}[leftmargin=*,nosep]
  \item \textbf{E8.} For every partition in \citetalias{paperA}'s
    Experiments E1--E6, compute the variance bracket and overlay
    against the entropy bracket. Validates
    \cref{cor:variance,thm:regression}. Expected output:
    \texttt{experiments/results/e8.json}.
  \item \textbf{E9.} Label-noise stress test on UCI Adult with
    $\eta_{\mathrm{noise}}\in\{0.05,0.10,0.20\}$. Compute both
    naive bracket on $\tilde f$ and the noise-corrected
    bracket per \cref{thm:noise}. Expected output:
    \texttt{experiments/results/e9.json}.
\end{itemize}
\status{Experiments deferred to Phase~3 of the master plan.
Skeletons live under \texttt{experiments/} with TODO headers.}

% ============================================================
\section{Discussion}\label{sec:disc}
% ============================================================

The $\bphi$-bracket meta-theorem clarifies that
\citetalias{paperA}'s elementary bracket is one face of a
larger object: every concave score functional with matching
loss gives a two-sided bracket on the partition-restricted
Bayes loss. The variance instance recovers the classical
Bayes--variance identity at equality; the Pinsker instance
sharpens it near boundary purities; the Shannon instance ships
to ML practitioners. Robustness extensions to label noise and
soft partitions show that the bracket is not an artefact of
deterministic, clean-label assumptions.

Open problems: (i) matching MAE lower bound for the regression
instance (\cref{thm:regression}); (ii) bracket consistency
under partition perturbation (Rand / VI stability, T5 in
\citetalias{paperA}'s audit); (iii) extending the meta-theorem
to multiclass with general concave score functionals on the
probability simplex.

% ============================================================
\bibliographystyle{tmlr}
\bibliography{main}

\end{document}
